---
title: "Lab 10. Dimensionality reduction. MDS"
output:
  html_document:
    df_print: paged
  pdf_document: default
editor_options:
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE)
```

```{r}
library(tidyverse)
library(ggplot2)
#library(ggmap)
#library(geosphere)
library(raster) # for pointDistance, distHaversine function to calculate distances using longitude and latitude coordinates
library(MASS) # for isoMDS
library(svglite) # to save svg files
```

## Multi-dimensional scaling

### 1. Location in poetry
The dataset `RNCpoetryLocation` presents the places where Russian verses where written. Source: Russian National Corpus Poetry database. 

* `Location` -- locations, mostly given by authors  
* `Coarse` -- less detailed and corrected data  
* `Region` -- region  
* `Decade` -- decade of text creation  
* `Freq` -- the number of texts created within the decade  
* `E` -- latude (North/South)  
* `N` -- longitude (East/West)

```{r}
geo0 <- read_tsv("https://raw.githubusercontent.com/LingData2019/LingData2020/master/data/RNCpoetryLocation.txt")
summary(geo0)

#geo <- distinct_all(geo0, vars(N, E), .keep_all = TRUE)
geo <- geo0 %>%
  group_by(N, E, Location=Coarse.1stMention, Reg=Region2) %>%
  summarize(Ndoc = sum(Freq)) %>%
  ungroup()
```

The dataframe `geo` contains only distinct Locations with their coordinates.

### 2. Distance matrix

A simple example demonstrates the function `pointDistance()` from the package `roster`.

Let us calculate the distances among `geo` locations. We use `pointDistance()` from `raster` package
```{r}
distmap <- pointDistance(geo[, c("E", "N")], lonlat=TRUE, allpairs=TRUE)
distmap[is.na(distmap)] = 0
distmap <- distmap + t(distmap)
distmap[1:4,1:4]
```

### 3. MDS

Fit MDS using the function `cmdscale()` (stats package).
```{r}
fit <- cmdscale(distmap, eig = TRUE)
dim(fit$points)
head(fit$points)
```

### 4. Plot the map
```{r}
geo2 <- data.frame(Dim1 = fit$points[,1], Dim2 = fit$points[,2], 
                   name = geo$Location, Ndoc = geo$Ndoc)
ggplot(geo2, aes(x=Dim2, y=Dim1, label=name)) +
  geom_point(aes(size=Ndoc), colour = 'darkgray', alpha = 0.7) +
  geom_text(aes(label=name), colour = 'orange', check_overlap = TRUE) 

# rotate coordinates
geo3 <- data.frame(lon = -1*fit$points[,1], lat = fit$points[,2],
                   name = geo$Location, Ndoc = geo$Ndoc, Reg = geo$Reg)
ggplot(geo3, aes(x=lon, y=lat, label=name)) +
  geom_point(aes(size=Ndoc, colour = factor(Reg)), alpha = 0.8) +
  geom_text(aes(label=name), colour = 'darkgray', size=2, 
            hjust=1, check_overlap = TRUE, vjust=1) +
  theme(panel.background = element_rect(fill = NA),
    panel.grid.major = element_line(colour = "grey90"),
    legend.position = "none")

# East European locations
ggplot(geo3, aes(x=lon, y=lat, label=name)) +
  geom_point(aes(size=Ndoc), colour = 'darkgray', alpha = 0.8) +
  xlim(-400000, 1000000) +
  ylim(-300000, 400000) +
  geom_text(aes(label=name), colour = 'blue', 
            check_overlap = TRUE, alpha = 0.7)
```

### Non-metric MDS
#### IsoMDS
```{r eval=FALSE}
# use distinct coordinates
fit.iso <- isoMDS(distmap)
geo.iso <- data.frame(name = geo$Location, lat = fit.iso$points[,1], lon = fit.iso$points[,2])
ggplot(geo.iso, aes(x=lon, y=lat, label=name)) +
  geom_point() +
  geom_text(aes(label=name), colour = 'darkgray', hjust=0, vjust=0, alpha = 0.5) 
```

```{r}
library(ggfortify)
row.names(distmap) <- geo$Location
fit.sam <- sammon(distmap)
kmeans_clust <- kmeans(fit.sam$points, 9)
str(kmeans_clust$cluster)
# rotate coordinates
geo4 <- data.frame(lon = -1*fit.sam$points[,1], lat = fit.sam$points[,2],
                   name = geo$Location, Ndoc = geo$Ndoc, 
                   Clust = as.factor(kmeans_clust$cluster))
ggplot(geo4, aes(x=lon, y=lat, label=name)) +
  geom_point(aes(size=Ndoc), colour='gray', alpha=0.8) +
  geom_text(aes(label=name), size=2, colour=geo4$Clust, 
            check_overlap=TRUE, alpha=0.7)

plot(fit.sam$points, type = "n", main="MDS with sammon() and clustered", xlab = "X-Dim", ylab="Y-Dim")
text(fit.sam$points, labels = rownames(distmap), col = kmeans_clust$cluster)
# simple view
autoplot(sammon(distmap), shape = FALSE, label.colour = 'blue', label.size = 3, label=TRUE, alpha = 0.5)
```

### Save as .svg file
```{r}
image <-
ggplot(geo3, aes(x=lon, y=lat, label=name)) +
  geom_point(aes(size=Ndoc, colour = factor(Reg)), alpha = 0.6) +
  xlim(-3000000, 5700000) +
  ylim(-2000000, 1700000) +
  geom_text(aes(label=name, size=Ndoc), 
          colour = 'grey20', alpha = 0.6, hjust = 'outward', vjust = 'outward') + # add check_overlap=TRUE in geom_text to avoid overlap 
  theme(panel.background = element_rect(fill = NA),
    panel.grid.major = element_line(colour = "grey90"),
    legend.position = "none")
image
  ggsave(file="test.svg", plot=image, width=20, height=8)
```

### Pre-training: four US cities
```{r}
df.cities1 <- data.frame(name = c("New York City", "Chicago", "Los Angeles", "Atlanta"),
                        lat  = c(       40.75170,  41.87440,      34.05420,  33.75280),
                        lon  = c(      -73.99420, -87.63940,    -118.24100, -84.39360))
distmap1 <- pointDistance(df.cities1[, c("lon", "lat")], lonlat=TRUE)
distmap1
distmap1[is.na(distmap1)] = 0
distmap1 <- as.dist(distmap1)
#distmap1 <- as.dist(t(distmap1)) 
#str(distmap1)
fit1 <- cmdscale(distmap1, eig = TRUE)
fit1$points
df.cities2 <- data.frame(name = df.cities1$name, 
                         lat = fit1$points[,1], 
                         lon = fit1$points[,2])
ggplot(df.cities2, aes(x=lon, y=lat, label=name)) +
  geom_point() +
  geom_text(aes(label=name), colour = 'orange', hjust=0, vjust=0)

# flip the map
df.cities3 <- data.frame(name = df.cities1$name, 
                         lat = -1*fit1$points[,1], 
                         lon = -1*fit1$points[,2])
ggplot(df.cities3, aes(y=lon, x=lat, label=name)) +
  geom_point() +
  geom_text(aes(label=name), colour = 'orange', hjust=0, vjust=0)
```

### Compare the picture
```{r}
library(rworldmap)
newmap <- getMap(resolution = "low")
plot(newmap, xlim = c(-10, 59), asp = 1)
points(geo$E, geo$N, col = "red", cex = .6)
```

### Other packages
```{r eval=FALSE}
library(ggfortify)
autoplot(eurodist)
autoplot(sammon(eurodist), shape = FALSE, label.colour = 'blue', label.size = 3)
autoplot(sammon(distmap), shape = FALSE, label.colour = 'blue', label.size = 3, label=geo$Coarse)
```

```{r eval=FALSE}
library(geosphere)
distHaversine(geo[, c("N", "E")]) / 1000 # Haversine distance in km
```
```{r eval=FALSE}
install.packages("ggmap")
library(ggmap)
map <- get_map(location = 'Europe', zoom = 4)
```

```{r eval=FALSE}
library(maps)
eu <- c("Austria", "Belgium", "Bulgaria", "Croatia", "Cyprus", "Czech Republic", 
        "Denmark", "Estonia", "Finland", "France", "Germany", "Greece", 
        "Hungary", "Ireland", "Italy", "Latvia", "Lithuania", "Luxembourg", 
        "Malta", "Netherlands", "Poland", "Portugal", "Romania", "Slovakia", 
        "Slovenia", "Spain", "Sweden", "United Kingdom")
europe <- map_data('world', region=eu)
library(ggplot2)
ggplot(europe, aes(x=long, y=lat, group=group)) +
  geom_polygon(fill="steelblue", colour="black") +
  xlim(-20, 40) + ylim(25,75)
```

```{r gene expression, eval=FALSE}
library(dynwrap)

url <- "https://www.ncbi.nlm.nih.gov/geo/download/?acc=GSE67310&format=file&file=GSE67310%5FiN%5Fdata%5Flog2FPKM%5Fannotated.txt.gz"
df <- read_tsv(url, col_types = cols(cell_name = "c", assignment = "c", experiment = "c", time_point = "c", .default = "d"))
expression <- df[, -c(1:5)] %>% as.matrix() %>% magrittr::set_rownames(df$cell_name)
cell_info <- df[, c(1:5)] %>% as.data.frame() %>% magrittr::set_rownames(df$cell_name) %>%
  rename(
    cell_id = cell_name,
    group_id = assignment
  ) %>%
  filter(group_id != "Fibroblast")

expression <- expression[cell_info$cell_id, expression %>% apply(2, sd) %>% order(decreasing = TRUE) %>% head(2000)]
counts <- 2^expression-1

fibroblast_reprogramming_treutlein <- wrap_data("id", rownames(expression)) %>%
  add_expression(counts, expression) %>%
  add_grouping(set_names(cell_info$group_id, cell_info$cell_id))
```

### Landmark MDS
```{r eval=FALSE}
set.seed(42)
install.packages("lmds")
library(lmds)
# compute distances between random landmarks and all data points
dist_landmarks <- select_landmarks(
  dataset$expression,
  distance_method = "pearson",
  num_landmarks = 150
)
dim(dist_landmarks)
# perform LMDS
dimred_lmds <- cmdscale_landmarks(dist_landmarks)
# now together
dimred_lmds2 <- lmds(
dataset$expression,
distance_method = "pearson",
num_landmarks = 150
)

# plot points
qplot(dimred_lmds[,1], dimred_lmds[,2], colour = dataset$grouping) +
theme_bw() +
labs(x = "Comp 1", y = "Comp 2", colour = "Group")

fit1 <- cmdscale(distmap1,eig=TRUE)
str(fit1)
autoplot(cmdscale(distmap1, eig = TRUE), label = TRUE, label.size = 3)
autoplot(fit1)
```
Source: https://www.r-bloggers.com/lmds-landmark-multi-dimensional-scaling/

Compare two lists with coordinates
```{r}
list1 <- data.frame(longitude = c(80.15998, 72.89125, 77.65032, 77.60599, 
                                  72.88120, 76.65460, 72.88232, 77.49186, 
                                  72.82228, 72.88871), 
                    latitude = c(12.90524, 19.08120, 12.97238, 12.90927, 
                                 19.08225, 12.81447, 19.08241, 13.00984,
                                 18.99347, 19.07990))
list2 <- data.frame(longitude = c(72.89537, 77.65094, 73.95325, 72.96746, 
                                  77.65058, 77.66715, 77.64214, 77.58415,
                                  77.76180, 76.65460), 
                    latitude = c(19.07726, 13.03902, 18.50330, 19.16764, 
                                 12.90871, 13.01693, 13.00954, 12.92079,
                                 13.02212, 12.81447), 
                    locality = c("A", "A", "B", "B", "C", "C", "C", "D", "D", "E"))
```
