---
title: "Lab6_anova"
output: html_document
date: "2023-01-31"
editor_options: 
  markdown: 
    wrap: 72
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r message=FALSE}
library(tidyverse) 
```

### Part 3: Universal linguistic hierarchies: a case of Modern Greek (Standard and Cypriot dialects)

Based on: Leivada, Evelina; Westergaard, Marit, 2019, Universal
linguistic hierarchies are not innately wired. PeerJ, v.7.
[link](https://www.ncbi.nlm.nih.gov/pmc/articles/PMC6679903/#fn-1).
Source of data: TROLLing repository: Leivada, Evelina; Westergaard,
Marit, 2019, "Replication Data for: Universal linguistic hierarchies are
not innately wired", <https://doi.org/10.18710/NTLLUF>, DataverseNO, V1

#### Constructions with two adjectives

In English, the order of two adjectives in phrases like:

     a big black bag # ok
    *a black big bag # unacceptable, ungrammatically ill-formed, or semantically anomalous

is powered by the semantic class of adjective (e.g. the `color`
adjective closer to the noun than the `size` adjective).

A syntactic hierarchy of closeness to the noun in Chomsky's Universal
Grammar suggests the following order and is claimed to be innate and
universal (= valid for all languages).

    Subjective Comment > Evidential > Size > Length
    > Height > Speed > Depth > Width > Temperature > Wetness > Age
    > Shape > Color > Nationality/Origin > Material 
    # (adapted from Scott, 2002: 114)

The goal of Leivada&Westergaard research is identify what happens when
people process orderings that either comply with the hierrarchy or
violate it.

#### Method

Experiment 1 140 neurotypical, adult speakers completed a timed forced
choice task that featured stimuli showing a combination of two
adjectives and a concrete noun (e.g., *I bought a square black table*).
Two types of responses were collected: (i) acceptability judgments on a
3-point Likert scale that featured the options 3: *correct*,\
2: *neither correct nor wrong*,\
1: *wrong*

and

(ii) reaction times (RT).

The task featured three conditions: 1. size adjective \> nationality
adjective, 2. color adjective \> shape adjective, 3. subjective comment
adjective \> material adjective. Each condition had two orders. In the
congruent order, the adjective pair was ordered in agreement with what
is traditionally accepted as dictated by the universal hierarchy. In the
incongruent order, the ordering was reversed, thus the hierarchy was
violated.

In the second experiment, 30 bidialectals (native speakers of Standard
and Cypriot Greek) were tested in both language varieties, 36
observations per participant, 18 for each variety.

#### Data

```{r}
suppressMessages(library(readr))
mono <- read_delim("https://dataverse.no/api/access/datafile/:persistentId?persistentId=doi:10.18710/NTLLUF/XAMMNB", delim = ";")

bidialect <- read_csv("https://dataverse.no/api/access/datafile/:persistentId?persistentId=doi:10.18710/NTLLUF/PQAKFW", header = F, sep = ";")
?read.csv
bidialect <- read_delim("https://dataverse.no/api/access/datafile/:persistentId?persistentId=doi:10.18710/NTLLUF/PQAKFW", delim=";")
```

see also [reading key for the
data](https://www.ncbi.nlm.nih.gov/pmc/articles/PMC6679903/bin/peerj-07-7438-s001.txt)

#### Data overview

### 1.1

```{r}

```

Create the following table using filter and by-group summary in
tidyverse.

Demographic information for monolinguals (experiment 1).

       Participants
    Gender     Male       66
               Female       74
    Education    Secondary  18
               Tertiary  122
    Handedness Right       126
               Left       14

### 1.2

```{r}

```

Create the same summary for experiment 2.

Demographic information for bidialectals (experiment 2).

    Participants
    Gender     Male       11
               Female       19
    Education    Secondary   6
               Tertiary   24
    Handedness Right        28
               Left        2

### 1.3

Create a plot that shows RT distribution in experiment 1 (all
participants and conditions taken together).

```{r}

```

Do reaction times follow a normal distribution? Or do they skewed and
have long left or right tails?

### 1.4

Normalise data applying the standard logarithm (RTlog = log10(RT))

```{r}

```

### 1.5

Create a plot similar to that in 1.3 that shows RTlog distribution.

```{r}

```

### 1.6

Filter out outliers: \* automatic responces below 600 ms (i.e., when a
button is pressed too fast, without allowing enough time for actual
consideration of the presented stimuli) \* usong ??3SD filter (SD -
standard deviation)

```{r}

```

### 1.7

Reproduce Figure 1 - Figure 7 from the paper using ggplot2.

### 1.8

Run ANOVA.

